### 2025-06-05 correlation.qmd
20250605 - correlation examples

The correlation coefficient ranges from −1.0 to +1.0.\index{correlation}
The correlation coefficient ($r$) tells you two things: (1) the direction (sign) of the association (positive or negative) and (2) the magnitude of the association.\index{correlation}
If the correlation coefficient is positive, the association is positive.\index{correlation}
If the correlation coefficient is negative, the association is negative.\index{correlation}
If the association is positive, as `X` increases, `Y` increases (or conversely, as `X` decreases, `Y` decreases).\index{correlation}
If the association is negative, as `X` increases, `Y` decreases (or conversely, as `X` decreases, `Y` increases).\index{correlation}
The smaller the absolute value of the correlation coefficient (i.e., the closer the $r$ value is to zero), the weaker the association and the flatter the slope of the best-fit line in a scatterplot.\index{correlation}
The larger the absolute value of the correlation coefficient (i.e., the closer the absolute value of the $r$ value is to one), the stronger the association and the steeper the slope of the best-fit line in a scatterplot.\index{correlation}
See @fig-rangeOfCorrelations for a range of different correlation coefficients and what some example data may look like for each direction and strength of association.\index{correlation}

@@

>>>>

The correlation coefficient ranges from −1.0 to +1.0.\index{correlation}
The correlation coefficient ($r$) tells you two things: (1) the direction (sign) of the association (positive or negative) and (2) the magnitude of the association.\index{correlation}
If the correlation coefficient is positive, the association is positive.\index{correlation}
If the correlation coefficient is negative, the association is negative.\index{correlation}
If the association is positive, as `X` increases, `Y` increases (or conversely, as `X` decreases, `Y` decreases).\index{correlation}
If the association is negative, as `X` increases, `Y` decreases (or conversely, as `X` decreases, `Y` increases).\index{correlation}
The smaller the absolute value of the correlation coefficient (i.e., the closer the $r$ value is to zero), the weaker the association and the flatter the slope of the best-fit line in a scatterplot.\index{correlation}
The larger the absolute value of the correlation coefficient (i.e., the closer the absolute value of the $r$ value is to one), the stronger the association.\index{correlation}
See @fig-rangeOfCorrelations for a range of different correlation coefficients and what some example data may look like for each direction and strength of association.\index{correlation}

@@

Also note that, although the the ([Pearson](#sec-correlationExamplesPearson)) correlation reflects the strength of the (linear) association between two variables, it does not reflect the slope of that association, as depicted in @fig-correlationExamples.

::: {#fig-correlationExamples}
![](images/correlationExamples.png){fig-alt="Several sets of ($x$, $y$) points, with the the ([Pearson](#sec-correlationExamplesPearson)) correlation coefficient of $x$ and $y$ for each set. The the ([Pearson](#sec-correlationExamplesPearson)) correlation reflects the strength and direction of a linear association (top row), but not the slope of that association (middle), nor many aspects of nonlinear associations (bottom). N.B.: the figure in the center has a slope of 0 but in that case the correlation coefficient is undefined because the variance of $y$ is zero. From <https://en.wikipedia.org/wiki/File:Correlation_examples2.svg>."}

Several sets of ($x$, $y$) points, with the the ([Pearson](#sec-correlationExamplesPearson)) correlation coefficient of $x$ and $y$ for each set. The the ([Pearson](#sec-correlationExamplesPearson)) correlation reflects the strength and direction of a linear association (top row), but not the slope of that association (middle), nor many aspects of nonlinear associations (bottom). N.B.: the figure in the center has a slope of 0 but in that case the correlation coefficient is undefined because the variance of $y$ is zero. From <https://en.wikipedia.org/wiki/File:Correlation_examples2.svg>.
:::
